Seasonality of Tidal Range
======
```{r setup, message=FALSE, echo=FALSE}
#options(width = 80, prompt = '>> ')
#setwd("/home/user/seasonality-of-tidal-range")  # linux
opts_chunk$set(out.width=650 , fig.align='center')
load("seasonalityTidalRangeData")
source("seasonalityTidalRange.R")
source("seasonalityTidalRange-1.R")
source("seasonalityTidalRange-3.R")
```

### Introduction

Theory informs us that tides are complex, influenced by geography of the undersea surface at the measuring station, gravitational forces of the Sun and Moon, and the oceanic basin a station is located in.<sup>[1], [2], [3]</sup> This analysis attempted to determine if there is a season in which the *tidal range* for Northern Hemisphere observing stations is greatest. The measure  *tidal range* is defined as *the difference in height between consecutive high and low waters*.<sup>[4]</sup> The questions we sought to answer: 

1. Is the *tidal range* in the Northern Hemisphere greatest during a solstice or an equinox?<sup>[5]</sup> If so, which one?
2. If a difference is detecte d, does the season-to-season change fit a linear or non-linear pattern? 

Theory says that the distance of a station from the Equator is just one factor in predicting a tide, and hence the *tidal range*. As a result, we didn't compare stations to each other without selecting a sufficiently large and random set of stations. Such a comparison was beyond the scope of this analysis. Instead, we aggregated water level observations to form a mean for a group of stations situated in a similar latitude, or region, and analyze the mean across seasons.

It might not actually be seasons that might explain variation in *tidal range* but the degrees of tilt of the Earth's axis. We used seasons here as a proxy for the degrees of tilt.

### Materials and methods
We performed secondary data analysis on water level data collected and curated by the National Oceanic and Atmospheric Administration (NOAA).<sup>[6]</sup> Because tide data is well kept, we do not expect variance in our data that is due to collection or curation methodologies. The datum we used was *MLLW* (mean lower low water), taken at six-minute intervals.

Data for 14 days before and 14 days after each solstice/equinox was be obtained. Each of the three data sets contained 29 days of data for each of the approximately 15 stations. This number of days, 29, was used to approximately match one, complete, 28-day lunar cycle, for each season.

The geodata for each station was converted from degrees, minutes, and seconds (DMS) to decimal using a freely available service<sup>[7]</sup>. Conversion is not complicated and we assume no variance was introduced into our data as a result of errors in the conversion. 

The map figure shows the locations and names of the stations selected for use in the analysis. The meaningful classification was by latitude. No regard was made to longitude; the hemispheric division in the figure was an artifact of the R technique used for coding locations. 

Our null hypothesis was that the *tidal range* for a region is the same during each of the four seasons. Since we do not have an expectation that *tidal range* does vary from season-to-season, this was the appropriate statement. We formed a null hypothesis and then tested the data. At the end of testing, we will show whether or not our station data provided evidence to support our null hypothesis.

The statement of our null hypothesis and testing procedure was as follows:

* $latex H_0: \mu_1 = \mu_2 = \mu_3  $
* $latex H_a: \mu_1 \neq \mu_2 \neq \mu_3 $
* Fit the data to a linear model and perform a one-way ANOVA to obtain test statistics
* If significance is determined, proceed to identify the groups that are different and attempt to explain why they are different

The respective values of $latex \mu $ in each test were the *mean of the difference between respective high and low tides for one 29 day period for one region for one season*. The *season* was the factor (categorical predictor value) that was used for the ANOVA.

An assumption we made is that that water levels for the autumnal equinox and vernal equinox are the same. As a result, we acquired and used data only the vernal equinox.

```{r, echo=FALSE, warning=FALSE, eval=TRUE, include=TRUE, message=FALSE, fig.height=5, fig.width=9, cache=TRUE}
require(maps)
require(RColorBrewer)
require(maptools)
require(classInt)
require(mapdata)
nclr <- 3
plotvar <- stationData$region
plotclr <- brewer.pal(nclr, "Set2")
class <- classIntervals(plotvar, nclr, style = "pretty")
colcode <- findColours(class, plotclr)
par(mfrow=c(1,2), 
	mar=c(0,0,1,0),
	xaxs = "i", 
	yaxs = "i", 
	bty = "n")
# eastern hemisphere
# dx <- 1.1^c()
map("world", 
	fill = TRUE, 
	col = "lightgrey", 
	bor = "lightgrey",
	xlim = c(90, 175), 
	ylim = c(0, 80))
title(main = "Eastern Hemisphere Stations",
	cex.main = 0.9)
text(stationData$long, stationData$lat, 
	stationData$friendlyName,
	pos = 2,
	cex = 0.9)
#
tempdf <- stationData
tempdf$region2 <- factor(tempdf$region, levels = c("low", "mid", "high"))
legend("topleft",
  legend = levels(tempdf$region2),
  fill = c(plotclr[1], plotclr[2], plotclr[3]),
  title = "latitude",
  box.lwd = 0,
  box.col = "white",
  bg = "white")
#
map.axes()
# western hemisphere
points(stationData$long, stationData$lat, 
	pch = 16, 
	col= colcode, 
	cex = 1.3)
map("world", 
	fill = TRUE, 
	col = "lightgrey", 
	bor = "lightgrey",
	xlim = c(-175, -90),
	ylim = c(0, 80))
title(main = "Western Hemisphere Stations",
	cex.main = 0.9)
points(stationData$long, stationData$lat, 
	pch = 16, 
	col= colcode, 
	cex = 1.3)
positions = c(4, 4, 4, 4, 2, 4, 2, 4, 4, 4, 2, 4) # from top to bottom
text(stationData$long, stationData$lat, 
	stationData$friendlyName,
	pos = positions,
	cex = 0.9)
map.axes()
title(main = "\nStation Names and Locations",
	outer = TRUE)
rm(tempdf, positions)
```
Station locations were carefully selected. Where we had personal knowledge of the location, we made a judgement on whether or not to include the station. Where we did not have personal experience with the location of a station, we instead relied on Google Maps to identify physical features that that might introduce variance to observations. For example, stations near the Columbia Bar, near the entrance to San Francisco Bay, or at the head of a fjord in Alaska, were disregarded as candidate sites. The ideal site was one that had unimpeded ocean water flow. For this reason, the Harvest Oil Platform situated off the coast of Lompoc, CA, was considered ideal. 

Counting a pairing of tides means that we start with the first high tide in our data for the station. It was possible that the  acquired data did not include the same count of high and low tides. We took this into account by avoided pairings of tides by taking calculating the range as the difference between the mean high water and the mean low water. (See the *Appendix* for acknowledgement of an error in this approach.)

### Results

We begin the review of our results by examining the data. The data shown in this table was aggregated from approximately 105,000 observations. The size of the resulting data set used in the linear models is only 45 records, 15 records per region. This table of descriptive statistics shows the data ordered by region and season:
```{r, echo=FALSE, warning=FALSE, eval=TRUE, include=TRUE, message=FALSE, results='asis'}
source("seasonalityTidalRange-4.R")
suppressPackageStartupMessages(require(googleVis))
chartVis <- gvisTable(summaryTable2, 'summary2', 
  options=list(height=350,
    showRowNumber = TRUE,
		width = 1000))
print(chartVis, tag='chart')
#plot(chartVis)
rm(chartVis)
```
To aid the reader in gaining some familiarity with the data, the columns can be interactively sorted by mouse clicks on the column header. 

Interesting patterns were revealed when we plotted the data in a boxplot of season by region. We saw quite clearly that in this data set, the median tidal range increased as we moved from low latitudes to higher latitudes. The interquartile range (IQR) of the high latitude tidal range was far wider than the range of either of the other two regions. Clearly, there was more variation in the water level readings obtained for the stations in Alaska. We discuss this characteristics of the data set further in our *Discussion* section. 
```{r echo=FALSE, warning=FALSE, eval=TRUE, include=TRUE, message=FALSE, fig.height=5, fig.width=9}
#load data, boxplot, summary tables, run your lm() model, summary(), plot(linear-model) to asses and #diagnose residuals, and the last line, anova(mylm)
tempdf <- summaryTable2
tempdf$region <- factor(tempdf$region, levels = c("low", "mid", "high"))
tempdf$season <- factor(tempdf$season, levels = c("winter", "spring", "summer"))
plotclr <- c("#66C2A5", "#FC8D62", "#8DA0CB")
colors = c(rep(plotclr[1], 3), rep(plotclr[2], 3), rep(plotclr[3], 3))
par(bty="n", mar=c(5, 4, 4, 0))
boxplot(tidalRangeMean ~ season * region, 
        data = tempdf,
        col = colors,
        ylab = "Tidal range (m)",
        xaxt = "n",
        boxwex = 0.40,
        main = "Tidal Range\nSeason * Latitude")
axis(1, at = 1:9, 
     labels = c("low\nwinter", "low\nspring", "low\nsum", 
        "mid\nwinter", "mid\nspring", "mid\nsum",
        "high\nwinter", "high\nspring", "high\nsum"),
     padj = 0.4)
legend("topleft",
       legend = levels(tempdf$region),
       fill = c(plotclr[1], plotclr[2], plotclr[3]),
       title = "latitude",
       box.lwd = 0,
       box.col = "white",
       bg = "white")
rm(tempdf)
```

```{r, echo=FALSE, warning=FALSE, eval=FALSE, include=FALSE, message=FALSE, results='asis', fig.height=5, fig.width=9}
#Here are the summary statistics for each station
source("seasonalityTidalRange-5.R")
require(googleVis)
chartVis <- gvisTable(summaryTable1, 'summary1', 
	options=list(height=250,
		width = 220))
print(chartVis, tag='chart')
#plot(chartVis)
rm(chartVis, dfTemp1, dfTemp2, dfTemp3)
```
The summary statistics from our ANOVA are summarized as follows:

* Region *low latitude*: F = 0.1204, with a p-value = 0.89.
* Region *mid latitude*: F = 0.6631, with a p-value = 0.53.
* Region *high latitude*: F = 0.0058, with a p-value = 0.99.

```{r, echo=FALSE, warning=FALSE, eval=TRUE, include=FALSE, message=FALSE}
lmLow <- with(subset(summaryTable2, region == "low"), lm(tidalRangeMean ~ season))
lmMid <- with(subset(summaryTable2, region == "mid"), lm(tidalRangeMean ~ season))
lmHigh <- with(subset(summaryTable2, region == "high"), lm(tidalRangeMean ~ season))
summary(lmLow)
summary(lmMid)
summary(lmHigh)
```
With one exception, the diagnostic plots, residuals against fitted values and qqnorm of residuals, appear to conform with the normality and similar variance across seasons. The exception is the high region that contains data from stations in Alaska. 
```{r, echo=FALSE, warning=FALSE, eval=FALSE, include=FALSE, message=FALSE}
par(mfrow=c(2,2))
plot(lmLow)
plot(lmMid)
```
```{r, echo=FALSE, warning=FALSE, eval=FALSE, include=FALSE, message=FALSE}
par(mfrow=c(2,2))
plot(lmHigh)
```
It's possible that the non-normal distribution for the high region is explained in part be explained by the stations we selected. The stations include Prudhoe Bay and Nome, AK, sites that are known to have frozen ocean water at the site for several months of the year. We are not educated on the impact sea ice has on the stations being able to record the water level. Further, the three remaining sites for the high region are located along the eastern shore of the Gulf of Alaska, an area known to have a large tidal range.
```{r, echo=FALSE, warning=FALSE, eval=TRUE, include=TRUE, message=FALSE}
anova(lmLow)
anova(lmMid)
anova(lmHigh)
```
Summarizing our results, the p-values for each of the three regions are very high, suggesting that it is not very unlikely to obtain the observed tidal ranges given the data available. We therfore <strong>fail to reject</strong> the null hyptohesis and conclude that there is not enough evidence to suggest that there is a relationship between the season and tidal range. 

### Discussion

In looking closely at the number of observations column, ```n```, in the table at the beginning of the *Results* section, we see that there are more high/low tide pairings than naturally occurs for a given period of 29 days. On average, we expect two high tides and two low tides per day for approximatley 58 complete pairings. Instead, for some stations, the number is many more than this. Obviously, the algorithm for building the high and low water levels is not sufficiently accurate. We discuss this in the *Appendix*.

Notwithstanding, the approach is valid and the results are sufficiently lacking any evidence for an effect that we are able to draw our conclusion that there is no effect of season on tidal range. 

The boxplot introduced in the *Results* section above shows an apparent pattern between latitude and tidal range. This pattern is clearly shown here:
```{r  echo=FALSE, warning=TRUE, eval=TRUE, include=TRUE, fig.height=3.5}
source("seasonalityTidalRange-5.R")
suppressPackageStartupMessages(require(ggplot2))
options(scipen=1)
ggplot(summaryTable1,
  aes(x = factor(season), y = mean, color = region, group = region)) + 
  	geom_line() +
		geom_point(size = 4, shape = 17) +
		ylab("Mean tidal range (m)") +
    scale_y_continuous(breaks = round(seq(0, 3.5, by = 0.5), 1)) +
    theme_bw() +
    xlab("Season") +
		theme(axis.title.y=element_text(size = 12)) +
		ggtitle("Mean Tidal Range by Season") + 
		scale_colour_brewer(palette = "Set2")
options(scipen=0)
```

We've included this figure for two reasons. First, visually inspecting this chart provides further confidence in our result that the season does not explain the variance in tidal range. But, and the primary reason for sharing this chart, is that it appears that there is a pattern of the tidal range increasing as the distance from the Equator increases, i.e., distance from the Equator would explain a significant portion of tidal range. However, we advise against studying this line of reasoning as there is not necessarily a correlation between latitude and tidal range. In the data selected for the high latitude region here, we know that three of the stations have particularly large tidal range compared to the readings obtained at the average station.

An interesting next step in investigating this data is to analyze the distribution of the times between tidal events. We know that natural phenomenon occur according to a distribution that we can undertand and can model. We would like to analyze this data set to see if the distribution of tidal events over time occur according to this distribution.

### Sources and references

1. Laplace theory of tides: http://en.wikipedia.org/wiki/Theory_of_tides
2. Tidal nodes: http://en.wikipedia.org/wiki/Amphidromic
3. Discussion of earth tides: http://en.wikipedia.org/wiki/Earth_tide
4. See *range of tide*, National Oceanic and Atmospheric Association (NOAA) glossary: http://tidesandcurrents.noaa.gov/glossary.html#R
5. Discussion on tidal datum: http://tidesandcurrents.noaa.gov/publications/tidal_datums_and_their_applications.pdf
6. Source data, retrieved by station location: http://tidesandcurrents.noaa.gov/
7. Conversion of DMS to decimal for station locations: http://andrew.hedges.name/experiments/convert_lat_long/

### Appendix

A portion of the custom code written for this analsysis is included below. As discussed in *Conclusions*, further work is necessary to control for fluctations in water level near maxima and minima that lead to falsely identifying turning points.

The approach taken in this analysis was to determine the local maxima and minima by comparing a set of observations later in time to the current observation. Then skipping 30, six-minute observations to avoid falsely identifying the next maxima or minima. The readings were at six-minute intervals so these 30 observations represented three hours of tidal action. We need to learn what the minimum time is between a maxima and minima in order to understand the ideal number of observations to bypass.

The algorithm is further explained in the comments for the code.

<pre><code># File name: seasonalityTidalRange.R
# Author: Phillip Burger
# Date: November 2013
# Project: seasonality-of-tidal-range
# This program: Builds high and low water vecctors used to determine tidal range.
# Description of the algorithm: We don't know if the first observation is a high
#   water, low water, on the way up, or on the way down. Definition of tidal
#   range is a high water minus the follow low water. Locate first high water
#   then the low water, and so on, until finished. Start by finding the trend. If
#   down, then skip first low water and find the following high water. If trend
#   is up, then start filling vector with this high water. At the end, take the 
#   elements in the shorter of the two vectors.

# Consumes water level data for one, 29-day period. Calcuates high and low 
#   water using the algorithm described in the file description.  
# Preconditions: Data for one period. Complete observations. 
#   Incomplete data removed prior to passing.  
# Postconditions: none 
# Parameters: waterLevel - data set contains water level observations.
# Return: Two element list. First element is high water, second is low water.

# Declare local variables. These are parameters used in identifying high and 
# low water.
skipNear <- 30
sampleNearMinMax <- 15
buildWaterLevels <- function(waterLevel) {
  highWater <- rep(0, 60)
  lowWater <- rep(0, 60)
  i <- 1  # index for high water
	j <- 1  # index for low water
	k <- firstHighWater(head(waterLevel, 250))  # get first high water
	highWater[i] <- waterLevel[k]
	i <- i + 1
	lookForHighWater <- FALSE  # find high tide if TRUE, find low tide if FALSE
	# traverse readings between ebb and flow, build tide levels at flood and ebb
	moreObs <- TRUE
	while (moreObs) {
		# look for next high tide, build high water.
		if (lookForHighWater) {
		  k <- k + skipNear  # skip near water to avoid false maxima
		  k <- k + tryCatch(getHighWater(waterLevel[k:length(waterLevel)]),
        error = function(err) {
          moreObs <- FALSE
        },
        finally = {})  # catch array out of bounds error
      highWater[i] <- waterLevel[k]
      i <- i + 1
		  lookForHighWater <- FALSE
		} else {
  		  k <- k + skipNear  # skip near water to avoid false minima
        k <- k + tryCatch(getLowWater(waterLevel[k:length(waterLevel)]),
          error = function(err) {
            moreObs <- FALSE
          },
          finally = {})  # catch array out of bounds error
        lowWater[j] <- waterLevel[k]
        j <- j + 1
        lookForHighWater <- TRUE  		  
		}
    if (k + skipNear > length(waterLevel)) {
			moreObs <- FALSE
		}
	}
  # number of elements with content be same length or highWater 1 longer
  if (i > j) {
    j <- j - 1
    highWater <- highWater[1:j]
    lowWater <- lowWater[1:j]
	} else {
    i <- i - 1
	  highWater <- highWater[1:i]
	  lowWater <- lowWater[1:i]     
	}
	return(list(highWater = highWater, lowWater = lowWater))	
}

# Helper function, find the first high tide for the observed period. This
#   function is called just one time. 
# Preconditions: 500 observations are passed in
# Postconditions: none.
# Parameters: waterLevel-data set containing first observations for the period.
# Return: Two element list. First element is high water, second is low water.
firstHighWater <- function(waterLevel) {
	k <- 1
	# simplest case, data starts with level increasing
  if (waterLevel[k] < mean(waterLevel[(k+1):(k+sampleNearMinMax)])) {
    k <- getHighWater(waterLevel)
  # less simple, data starts with level decreasing 
  } else {
    k <- getLowWater(waterLevel)  # identity the first minima
    k <- k + getHighWater(waterLevel[(k+1):length(waterLevel)])  # then identify fresh maxima, not dirty one if data starts with tredning down.
  }
	return(k)
}

# Helper function, find the first high water level in the vector
# Parameters: waterLevel-data set containing first observations for the period.
# Return: Index location in waterLevel marking the local maxima
getHighWater <- function(waterLevel) {
  k <- 1
  trendingHigher <- TRUE
  while (trendingHigher) {
    if (waterLevel[k] < mean(waterLevel[(k+1):(k+sampleNearMinMax)])) {
      k <- k + 1
    } else {
      k <- (k - 1) + which.max(waterLevel[k:(k+1+sampleNearMinMax)])
      trendingHigher <- FALSE  # found high water, return k
    }
  }
  return(k)
}

# Helper function, find the first low water level in the vector
# Parameters: waterLevel-data set containing first observations for the period.
# Return: Index location in waterLevel marking the first local minima
getLowWater <- function(waterLevel) {
  k <- 1
  trendingLower <- TRUE
  while (trendingLower) {
    if (waterLevel[k] > mean(waterLevel[(k+1):(k+sampleNearMinMax)])) {
      k <- k + 1
    } else {
      k <- (k - 1) + which.min(waterLevel[k:(k+1+sampleNearMinMax)])
      trendingLower <- FALSE  # found low water, return k
    }
  }
  return(k)
}
</code></pre>

The data and code are freely available in a Github repository under user [sculpturearts](https://github.com/sculpturearts/seasonality-of-tidal-range).